---
title: "Getting started with lapidary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lapidary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(lapidary)
```

`lapidary` has three layers: a **canonical data model**, **analysis
primitives**, and a **shared visual style**. This vignette walks through them
with the packaged sample of GEMS-GER (`?gems_ger_sample`), so nothing here
needs the full download.

## The data model

Every source is normalised to two shapes:

* `gwl_ts` — a tidy long tibble, one row per well × timestamp
  (`well_id`, `date`, `gwl`, `variable`, `source`, optional `gwl_flag`);
* `gwl_wells` — an `sf` POINT layer of well metadata in EPSG:25832.

```{r}
data(gems_ger_sample)
gems_ger_sample

validate_gwl_ts(gems_ger_sample)
```

`variable` records *what* `gwl` is (here `"gwl_m_asl"`, metres above sea
level), so series with different datums never get mixed silently.

To ingest the real dataset:

```{r eval = FALSE}
lap_gems_ger_download()                    # ~290 MB, once
lap_gems_ger_build_parquet(meteo = FALSE)  # CSV -> Parquet, once
gwl <- lap_read_gems_ger(date_range = c("2000-01-01", "2020-12-31"))
```

Adding another source means writing a reader that returns these shapes;
`lap_read_correctiv()` is a worked second example.

## Analysis primitives

```{r}
enriched <- gems_ger_sample |>
  lap_use_water_year(start_month = 11) |>
  lap_add_reference_period(periods = list(Z1 = c(1991, 2020)))

annual <- lap_summarise_wells(enriched, by = c(well_id, year))
head(annual)
```

Trends use the Theil–Sen slope with a Mann–Kendall significance test:

```{r}
trends <- lap_gw_trend(annual, value = mean_gwl, time = year)
head(trends[order(trends$slope), c("well_id", "slope", "p_value", "direction")])
```

Normalisation makes wells comparable:

```{r}
sgi <- lap_normalise_gwl(gems_ger_sample, method = "sgi")
range(sgi$gwl_norm, na.rm = TRUE)
```

## Spatial aggregation

```{r}
data(germany_hex_sample)
germany_hex_sample
```

`germany_hex_sample` was built with `lap_aggregate_to_hex()` from the sample
wells; with the full dataset the hexagons are densely populated.

## Style

```{r fig.height = 5}
library(ggplot2)

ggplot(subset(germany_hex_sample, n_wells > 0)) +
  geom_sf(aes(fill = mean_gwl), colour = "white", linewidth = 0.1) +
  scale_fill_lapidary_c("magnitude") +
  theme_lapidary(variant = "light") +
  labs(title = lap_tr("app_title"), fill = lap_tr("groundwater_level"))
```

Switch `variant = "dark"`, set `options(lapidary.lang = "de")`, and use
`ggsave_lapidary(preset = "a1")` for print output. See
`vignette("style", package = "lapidary")`.
