---
title: "The lapidary style scheme"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The lapidary style scheme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(lapidary)
library(ggplot2)
```

The visual style is one system with a light and a dark variant, driven by a
single set of **design tokens**. Plot code reads tokens through
`lap_tokens()` rather than hard-coding values, so a restyle happens in one
place (see `docs/adr/0003`).

## Tokens

```{r}
tk <- lap_tokens("light")
str(tk, max.level = 1)
tk$colour$recharge
tk$colour$discharge
```

Type sizes are multipliers of a `base_size`, which is how the *same* plot
scales from web to A0.

## Theme and scales

The continuous `scale_*_lapidary_c()` scales are **binned** at pretty round
breaks and drawn as a long, thin colour-steps bar; the legend title is
prettified from the mapped variable (see `lap_prettify_label()`) unless you set
`name`/`labs()`; and `lap_na_guide()` adds a "no data" key for `NA` regions such
as empty hexes. With `scale_fill_lapidary_c(range = TRUE)` — or
`options(lapidary.scale_range = TRUE)` — a mapped indicator column's theoretical
range is appended to the title (`"Trend slope  (-Inf, Inf)"`).

```{r}
data(germany_hex_sample)

p <- ggplot(germany_hex_sample) +
  geom_sf(aes(fill = mean_gwl), colour = "white", linewidth = 0.1) +
  scale_fill_lapidary_c("magnitude") +
  lap_na_guide() +
  labs(title = lap_tr("app_title"), fill = lap_tr("groundwater_level"))

p + theme_lapidary("light")
```

```{r}
p + theme_lapidary("dark")
```

A handful of extreme values can flatten the colour range for everywhere else.
`robust = TRUE` clips the limits to the 2nd and 98th percentiles (pass a number
like `0.95`, or `c(0.05, 0.95)`, to choose them; `options(lapidary.scale_robust
= TRUE)` sets the default). Clipped values take the end colour and the binned
legend marks that end with `≤` / `≥`:

```{r}
set.seed(1)
hex_spike <- germany_hex_sample
hex_spike$mean_gwl[sample(which(!is.na(hex_spike$mean_gwl)), 5)] <- 5000

ggplot(hex_spike) +
  geom_sf(aes(fill = mean_gwl), colour = "white", linewidth = 0.1) +
  scale_fill_lapidary_c("magnitude", robust = TRUE) +
  lap_na_guide() +
  theme_lapidary("light")
```

Palette roles decouple intent from the underlying scico palette:

```{r}
lap_pal_roles()
```

* `months` – cyclic, for month-of-year / phase
* `magnitude` / `magnitude_dark` – sequential
* `density` – sequential greys, for counts
* `anomaly` – divergent, for departures from a reference

## Fonts

`lap_fonts()` registers Oleo Script (titles) and Dosis (body) and enables
`showtext`. If the fonts or packages are missing it is a no-op and the theme
falls back to `serif` / `sans`, so headless rendering never breaks.

```{r eval = FALSE}
lap_fonts()
```

## Output presets

```{r}
lap_preset_names()
lap_preset("a1")
```

`ggsave_lapidary()` applies a preset's width/height/dpi *and* sets
`showtext`'s DPI to match, which is the difference between crisp and blurry
poster text.

```{r eval = FALSE}
ggsave_lapidary(p + theme_lapidary("light", base_size = lap_preset("a1")$base_size),
  "groundwater_a1.png",
  preset = "a1"
)
```

## Language

```{r}
lap_tr("groundwater_level", "en")
lap_tr("groundwater_level", "de")
```

Set a session default with `options(lapidary.lang = "de")` or the
`LAPIDARY_LANG` environment variable.
